Reasoning about a Capability Machine with Local Capabilities

### 2017-06-15 popl-2018/paper.tex
rewrote a bit of Worlds subsection to fit better with new introduction

\subsection{Formalizing the guarantees of the capability machine}
What differentiates a capability machine from a more standard assembly language
is that we can bound the authority of an executing block of code, based solely
on the capabilities it has access to. Specifically, it does not matter which
instructions are actually executed, i.e., the bound also applies to untrusted
adversary code that has not been inspected or modified in any way.

@@

\paragraph{Step-indexing}
More generally, we want to define that a read capability for memory range
$[b,e]$ is safe, if the world guarantees that the words at those addresses are
themselves safe. However, this definition is cyclic: what if the world
guarantees that the memory at address $a$ will contain a read capability for
address $a$? The definition then just says that a read capability for address
$a$ is safe iff the same read capability for address $a$ is safe. This form of
cyclic reasoning is related to similar challenges in languages with recursive
types or higher-order ML-style references, and a standard solution is
to use step-indexing: Essentially, the cycle is broken by defining safety up to a
certain number of interaction steps. All words will be considered
safe up to $0$ steps (since if there no interactions, nothing unsafe
can happen), and, for example, a read capability will be safe up to $n$
steps if the world guarantees that the words at the corresponding addresses are
safe up to $n - 1$ steps. We can then prove that the above read capability for
address $a$ is safe up to any number of steps.

@@

\paragraph{Invariants and Recursive Worlds}
So worlds group a set of memory invariants, but how are they actually
defined formally? We represent each invariant by a region
$r \in \Regions$. We will see later that regions contain state
machines to support a notion of evolvable invariants, but in every
state, they also contain a predicate $H$ that defines which memory
segments are acceptable in the current state of the invariant.
Unfortunately, it is not enough to just take
$H \in \powerset{\MemSegments}$, because sometimes the invariant may
itself be world-dependent. For example, we may want to express
invariants like ``the memory at address $50$ contains a value that is
safe in the current world''. As explained, worlds may evolve, and the
set of safe values may grow in future worlds, and therefore we need to
index $H$ over worlds, i.e., take
$H \in \Worlds \rightarrow \powerset{\MemSegments}$. We then end up
with worlds containing regions and regions that are world-indexed,
i.e., the set of worlds must be recursively defined. We will see how such a recursive
definition can be accommodated using techniques from the
literature (essentially an advanced application of step-indexing).

\paragraph{Local capabilities}
A final thing we explain here is how our results take into account local
capabilities and their special treatment by the hardware. Normally, when we
invoke an untrusted piece of code and provide it with certain global
capabilities, it may have stored those capabilities in memory and we will only
be able to reinvoke the code if we can guarantee that those values are still
valid. Formally, worlds represent the invariants that the global capabilities'
safety relies on and the reinvocation is only safe in future worlds, where the
invariants are respected.

However, if we provide the adversary with local capabilities in that first
invocation, then the situation is a bit different. The adversary has no way to
store these local capabilities, so if we make sure that there are also no local
capabilities in the register file for the second invocation (including the
capability being invoked), then the adversary cannot use them any more.%
\footnote{We are ignoring store-local capabilities in this discussion. If the
  adversary does have access to store-local capabilities in the first and second
  invocation, then the memory they address must entirely be cleared before the
  second invocation in order for the reinvocation to remain safe.} Therefore, we
can allow the second invocation to happen in \emph{private} future worlds ($W'
\futurestr W$), in which global safe capabilities remain safe, but local ones
don't. This private future world relation is more liberal than the standard,
public one ($W' \futurewk W$, in which \emph{all} safe capabilities remain
safe). Concretely, worlds may contain \emph{temporary} regions, representing
invariants that only local capabilties may rely on for their safety, and which
may be revoked (disabled) in private future worlds.

@@

\subsection{Worlds}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{An illustration of local/global capabilities and temporary/permanent
    regions. The table represents a memory and the colored fields are regions
    governing a memory range. Global capabilities cannot depend on temporary
    regions.}
  \label{fig:cap-world}
\end{figure}
% \dominique{Perhaps we could first try to convey the intuition of a permanent
%   world with a temporary overlay, with global capabilities ignoring the overlay.
%   Perhaps we could add a depiction of the world to Figure 5?}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}
Formally, a world is a finite map from region names, modeled as natural numbers, to regions.
We have three types of regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe a protocol for the region's memory,
similarly to what has been used in logical relations for high-level
languages~\citep{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked at certain points during execution, on the condition that we
clear local capabilities when this happens. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
A region defines what its part of memory can legally look like. However,
validity of the memory's contents may itself depend on the state of other parts
of memory. To accommodate this, worlds need to be defined recursively: Worlds
describe memory, which can contain capabilities, whose safety is
world-dependent. Recursive worlds are a common requirement in Kripke models and
the following lemma uses the method of
\citet{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the
aforementioned state transition system where $\Rels$ is a set of pairs
of relations corresponding to the public and private transition, and
$\States$ is a sufficiently large set of states containing at least the
states we use. The last part of the temporary and permanent regions is
a state interpretation function that determines what memory segments
the region permits in each state of its states. The different
monotonicity requirements in the two interpretation functions reflects
how permanent regions rely only on permanent protocols whereas
temporary regions can rely on both temporary and permanent protocols.
$\UPred{\HeapSegments}$ is a step-indexed, downwards closed predicate
on memory segments.

>>>>

\subsection{Formalizing the guarantees of the capability machine}
\label{sec:formalizing-guarantees}
What differentiates a capability machine from a more standard assembly language
is that we can bound the authority of an executing block of code, based solely
on the capabilities it has access to. Specifically, it does not matter which
instructions are actually executed, i.e., the bound also applies to untrusted
adversary code that has not been inspected or modified in any way.

@@

\paragraph{Step-indexing}
More generally, we want to define that a read capability for memory range
$[b,e]$ is safe, if the world guarantees that the words at those addresses are
themselves safe. However, this definition is cyclic: what if the world
guarantees that the memory at address $a$ will contain a read capability for
address $a$? The definition then just says that a read capability for address
$a$ is safe iff the same read capability for address $a$ is safe. This form of
cyclic reasoning is related to similar challenges in languages with recursive
types or higher-order ML-style references, and a standard solution is
to use step-indexing: Essentially, the cycle is broken by defining safety up to a
certain number of interaction steps. All words will be considered
safe up to $0$ steps (since if there is no interaction, nothing unsafe
can happen), and, for example, a read capability will be safe up to $n$
steps if the world guarantees that the words at the corresponding addresses are
safe up to $n - 1$ steps. We can then prove that the above read capability for
address $a$ is safe up to any number of steps.

@@

\paragraph{Invariants and Recursive Worlds}
So worlds group a set of memory invariants, but how are they actually
defined formally? We represent each invariant by a region
$r \in \Regions$. We will see later that regions contain state
machines to support a notion of evolvable invariants, but in every
state, they also contain a predicate $H$ that defines which memory
segments are acceptable in the current state of the invariant.
Unfortunately, it is not enough to just take
$H \in \powerset{\MemSegments}$, because sometimes the invariant may
itself be world-dependent. For example, we may want to express
invariants like ``the memory at address $50$ contains a value that is
safe in the current world''. As explained, worlds may evolve, and the
set of safe values may grow in future worlds, and therefore we need to
index $H$ over worlds, i.e., take
$H \in \Worlds \rightarrow \powerset{\MemSegments}$. We then end up
with worlds containing regions with world-indexed predicates,
i.e., the set of worlds must be recursively defined. We will see how such a recursive
definition can be accommodated using techniques from the
literature (essentially an advanced application of step-indexing).

\paragraph{Local capabilities}
A final thing we explain here is how our results take into account local
capabilities and their special treatment by the hardware. Normally, when we
invoke an untrusted piece of code and provide it with certain global
capabilities, it may have stored those capabilities in memory and we will only
be able to reinvoke the code if we can guarantee that those values are still
valid. Formally, worlds will represent the invariants that the global
capabilities' safety relies on and reinvocation will only be possible in future
worlds.

However, if we provide the adversary with local capabilities in the first
invocation, then the situation is a bit different. Since the adversary has no
way to store local capabilities, we can allow the second invocation to happen in
\emph{private} future worlds ($W' \futurestr W$), in which global capabilities
remain safe, but local ones don't. This private future world relation is more
liberal than the public one ($W' \futurewk W$, in which \emph{all} safe
capabilities remain safe). In particular, worlds may contain \emph{temporary}
regions, representing invariants that only local capabilties may rely on for
their safety, and which may be revoked (disabled) in private future worlds.

@@

\subsection{Worlds}
\begin{figure}
  \centering
  \includegraphics{w11}
  \caption{An illustration of local/global capabilities and temporary/permanent
    regions. The table represents a memory and the colored fields are regions
    governing a memory range. Global capabilities cannot depend on temporary
    regions.}
  \label{fig:cap-world}
\end{figure}
% \dominique{Perhaps we could first try to convey the intuition of a permanent
%   world with a temporary overlay, with global capabilities ignoring the overlay.
%   Perhaps we could add a depiction of the world to Figure 5?}
%\todo[inline]{Find a good place to describe monotonicity of
%interpretation function.}
% \lau{this may not be an important enough point to include it.}

As describe above, worlds are a collection of invariants. Formally, a world is a
finite map from region names, modeled as natural numbers, to regions that each
correspond to an invariant of part of the memory.  We have three types of
regions: \emph{permanent}, \emph{temporary}, and \emph{revoked}.
%$\perma$, $\temp$, and $\revoked$. The
%$\perma$ and $\temp$ 
Each permanent and temporary region contains a state transition system, with
public and private transitions, to describe how the invariants are allowed to
change over time. In other words, they are protocols for the region's
memory. These are similar to what has been used in logical relations for
high-level
languages~\citep{Ahmed:popl09,Dreyer:jfp12,Devriese:2016ObjCap}. Protocols
imposed by permanent regions stay in place indefinitely. Any capability, local
or global, can depend on these protocols. Protocols imposed by temporary regions
can be revoked at certain points during execution, on the condition that we
clear local capabilities when this happens. This means that local capabilities
can safely depend on the protocols imposed by temporary regions, but global
capabilities cannot, since a global capability may outlive a temporary region
that is revoked. This is illustrated in \figurename~\ref{fig:cap-world}.

@@

%\dominique{``end up being defined recursively'' does not sound very insightful}
% A region defines what its part of memory can legally look like. However,
% validity of the memory's contents may itself depend on the state of other parts
% of memory. To accommodate this, worlds need to be defined recursively: Worlds
% describe memory, which can contain capabilities, whose safety is
% world-dependent. 
As described in Section~\ref{sec:formalizing-guarantees}, the safety of a world
depends on the safety of its regions which in turn may depend on the safety of
the values it contain which depend on the safety of a world. In other words, we
need recursive worlds. Recursive worlds are a common requirement in Kripke
models and the following lemma uses the method of
\citet{Birkedal:2011:SKM:1926385.1926401,Birkedal:tutorial-notes} for
constructing them. The formulation of the lemma is technical, so we recommend
that non-expert readers ignore the technicalities and accept that there exists a
set of worlds $\Wor$ and two relations $\futurestr$ and $\futurewk$ satisfying
the (recursive) equations in the theorem (where the $\blater$ operator can be
safely ignored).
% Short technical explanation
% The recursiveness of the worlds gives rise to a recurisve domain
% equation. Theorem~\ref{thm:world-existence} gives the solution to this
% recursive domain equation.
% \dominique{it is not explained what cofes are?}
% \lau{I added what cofe means but no formal definition (I think this goes under the category technicalities)}
\begin{theorem}\label{thm:world-existence}
  There exists a \cofe{} (complete ordered family of equivalences) $\Wor$ and preorders $\futurestr$ and
  $\futurewk$ such that $(\Wor,\futurestr)$ and $(\Wor,\futurewk)$ are
  preordered \cofes{}, and there exists an isomorphism $\xi$ such that
  \begin{align*}
      \xi \; : \; & \Wor \cong \blater (\nats \finparfun \Regions)\\
      \Regions  = & \; \{\revoked\} \uplus \\
               & \; \{\temp\} \times \States \times \Rels \times (\States \fun (\Wor \monwknefun \UPred{\HeapSegments}))\uplus \\
               & \; \{\perma\} \times \States \times \Rels \times (\States \fun (\Wor \monstrnefun \UPred{\HeapSegments}))
    \end{align*}
  and for $W, W' \in \Wor$
  \begin{align*}
    W' \futurestr W & \Leftrightarrow \xi(W') \futurestr \xi(W)   \\
    W' \futurewk W & \Leftrightarrow \xi(W') \futurewk \xi(W)
  \end{align*}
\end{theorem}
In the above theorem, $\States \times \Rels$ corresponds to the
aforementioned state transition system where $\Rels$ is a set of pairs
of relations corresponding to the public and private transition, and
$\States$ is a sufficiently large set of states containing at least the
states we use. The last part of the temporary and permanent regions is
a state interpretation function that determines what memory segments
the region permits in each state of its states. The different
monotonicity requirements in the two interpretation functions reflects
how permanent regions rely only on permanent protocols whereas
temporary regions can rely on both temporary and permanent protocols.
$\UPred{\HeapSegments}$ is a step-indexed, downwards closed predicate
on memory segments.
